DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-12-15 2015CompleteDerivation/completeDerivation.tex
First pass at the introduction

We work strives as much as possible to be philosophically consistent, physically meaningful and mathematically precise.
\end{abstract}
\maketitle

\section{Status}

@@

@@

The work is organized into:
\begin{description}
  \item[Assumptions] these characterize the physical system we are studying and constitute the premise of our discussion. A \textbf{rationale} follows each assumption, which uses physical and sometimes philosophical arguments to motivate why (or why not) such an assumption makes sense (in a particular case).
  \item[Propositions] these characterize the mathematical properties of the physical concepts we are studying. A \textbf{justification} follows each proposition to show why such characterization follows from the physical properties of the system. While the style of a justification is formal, it is not a pure mathematical proof as it necessarily has to mix physical and mathematical arguments.

  Before or after each proposition there is a \textbf{discussion} that conceptually explains the motivation and the result in a less formal, more approachable way.

@@

\section{On studying a physical system}

@@

\section{States and state space}

@@

\section{Time dependent evolution}

@@

\section{Kinematics and minimal materials}

@@

\section{Prelude to quantum}

@@

\section{Irreducible systems}

@@

\section{Discussion and further work}

@@

We have seen that much of fundamental physics can be derived from a few simple physical assumption (and lots of hard work). In doing so we obtained a more unified picture of the different branches of fundamental physics centered around the idea of deterministic and reversible evolution of distributions of material. The main physical concepts can be summarized by the following points:

@@

\item The state space is not just a collection of states but also captures its physical properties such as physical distinguishability (represented by a topology), the ability to count states (represented by a measure, symplectic form or metric tensor) or decomposability (represented by a vector space structure).

@@

\item Classical Hamiltonian mechanics describes the evolution of a homogeneous material for which the state is equivalent to the states of all its infinitesimal parts, the classical particles. The state space for classical particles is the symplectic manifold $(T^*\mathcal{Q}, \omega)$; deterministic and reversible evolution is a symplectomorphism (i.e. a canonical transformation).

\item If the classical particle states are fully characterized by their motion in space, then the states also evolve according to Lagrangian mechanics and the motion is that of a massive particle under scalar/vector potential forces.

@@

\item Quantum (Hamiltonian) mechanics describes the evolution of a homogeneous material for which the state can only be reduced up to finite parts, the quantum particles, whose internal motion is undetermined. The state space for quantum particles is an inner vector product space $(\Psi, \langle \cdot , \cdot \rangle)$; deterministic and reversible evolution is a unitary transformation.
\end{itemize}

While we are sure that some mathematical details can be improved, the feel that the overall ``conceptual castle" is solid. The starting concepts are simple, the conceptual mapping of those concepts into math is often straight forward and it often gave intuitive ideas that connects to different branches of math and physics: it cannot be a simple coincidence.

While the scope of this work is limited to re-founding and re-understanding the already known physical theories, it is our hope that this insight may also lead to new ideas to address some of the open problems in the foundations of physics.

>>>>

This work strives as much as possible to be philosophically consistent, physically meaningful and mathematically precise.
\end{abstract}
\maketitle

\textbf{\emph{Status}}

@@

\section{Introduction}

While some branches of physics (e.g. Newtonian mechanics, thermodynamics, special relativity) are derived from physical assumptions (e.g. Newton's laws, laws of thermodynamics, invariance of the speed of light), others start by simply setting a mathematical framework. It is the case for classical Lagrangian mechanics, where we start by assuming the existence of a Lagrangian function whose path integral is minimized along the motion. It is the case for classical Hamiltonian mechanics, where we start by assuming the state is described by conjugate pairs $(q^i, p_i)$ and evolves according to Hamilton's equations. It is the case for quantum mechanics, where the state space is a complex inner product vector space and the state evolves according to Schroedinger's equations. That is, at the very core of many of our physical theories we have mathematical definitions instead of physical concepts.

While we have grown accustomed to this state of affair, it is nonetheless troublesome as it has problematic consequences. Suppose you have a system in front of you, how can you tell whether it's Hamiltonian? Or classical? We know that these theories have different range of applicability. To describe a cannonball we should be using classical mechanics while for an electron we should be using quantum mechanics. Not all Hamiltonians admit a corresponding Lagrangian. How do we pick?

If all we have are mathematical definitions, the only answer is that it is a Hamiltonian system if it obeys Hamiltonian mechanics. While many are indeed satisfied by such answer, one has to admit that it begs the question. We do not have a clear idea of what an Hamiltonian system is and how it different from a Lagrangian one. Or what makes a system classical and how it is different form a quantum system.

We started this work with the unpretentious goal of exploring these questions. The idea was to create a math-to-physics dictionary so that each mathematical concept was clearly mapped to a crisp physical concept. But to make sure that such dictionary was complete, we had to make sure that all the relevant mathematical framework could be recovered. This lead to the mission of re-deriving as much as possible starting from physical assumptions.

This approach pushed us further and further back to the basic definitions. For example, if we truly wanted to justify why the state space of Hamiltonian mechanics is charted by conjugate variables $q^i$ and $p_i$, we had to justify why it was a manifold as well, which in turn meant we needed to justify why it was a topological space. But when we did find physical concepts that mapped to these more basic mathematical definitions, we realized that this link was valid for a far wider range of cases: the concepts were more general than we expected.

To our surprise, this pushed us to bring in other areas that were not within our horizon. The same link that allowed to justify Hamiltonian dynamics also gave us relativistic motion on the extended phase space. The same link that allowed to justify Lagrangian mechanics also gave us constrained the motion to free particles under scalar/vector potential forces. Since this happened repeatedly, it lead to think that, maybe, the picture we were developing was not simply ``a nicer way to think about things", but a fundamental framework to understand and link different parts of basic science and mathematics.

%It felt like, when trying to assemble a jigsaw puzzle, you realize that the reason you can't find a piece is because you assembled the previous parts incorrectly and you end with a larger part that fits together a lot better.

These unexpected connections lead to the realization that sometimes different mathematical concepts are expression of the same physical concept. That is: by putting physics first we can achieve a sort of unification of ideas that is not possible by math alone.

The goal of this particular work is to report on this progress. The claim is that the whole architectural framework is solid enough to be convinced that it is possible to derive classical/quantum Hamiltonian/Lagrangian mechanics starting from a few physical principle. This is not just an exercise in style: this approach allows to make intuitive connections between different areas of physics and mathematics.

The claim is that, starting from a few simple physical assumptions, we can derive classical/quantum Hamiltonian/Lagrangian mechanics. The claim is not that we have achieved perfection and that every proposition and justification cannot be improved. The claim is, simply, that it \emph{can} be done, and that doing so provides better insight and intuition that clarifies the physics being described and that allows to strengthen connections between different areas of math and physics. We will see that sometimes different mathematical concepts are the expression of one physical concept. Therefore, by putting physics, first we can achieve a sort of unification of ideas that is not possible by math alone.

While trying to re-derive classical, quantum and (to some extent) relativistic mechanics may seem too broad a scope, it is necessary to ensure that the concepts we develop are truly general, that they can be applied to many different cases. If we could derive classical Hamiltonian mechanics on assumptions that are inapplicable to the quantum case, what good would it do? Quantum system are Hamiltonian systems, so clearly we haven't understood what makes a system Hamiltonian. The ability to apply the same physical concepts far and wide is what gives us confidence that these are truly general ideas and they capture the physical aspects we are describing.

As brief overview, there are four assumptions we will consider. Assuming deterministic and reversible evolution will give us a dynamical system: a state and a law of evolution. Additionally assuming infinitesimal reducibility, studying the whole system is equivalent to studying its infinitesimal parts, gives us Hamiltonian mechanics. Additionally assuming kinematic equivalence, trajectories in space tell us everything about the system, gives us Lagrangian mechanics and constrains the motion to massive particles under scalar/vector potential forces. Relativistic motion arises when studying time dependent laws of evolution with no further assumption. Alternatively, as the second assumption we can take irreducibility, studying the whole systems tells us nothing about its parts, which leads to quantum (Hamiltonian) mechanics.

In section \ref{sec:fundamental_model} we will describe in general what it means to study a physical system and how the assumption of determinism and reversibility plays into it. We will develop an abstract conceptual model that we will use when discussing the later assumptions.

In section \ref{sec:dynamical_systems} we will show how the physical concept of physical distinguishability maps to the mathematical concept of topological spaces. Deterministic and reversible is an invertible continuous map (i.e. a homeomorphism) as it needs to preserve what is physically distinguishable.

In section \ref{sec:reducibility} we will introduce the notion of a classical material, one for which we assume infinitesimal reducibility. The state of an arbitrary amount of material is a distribution over the states of the infinitesimal parts, the classical particles. As such distribution must be invariant under the arbitrary choice of units, for each variable $q^i$ that defines a unit there is a corresponding variable $k^i$ defined on the inverse of such unit. Classical particle states can be identified by conjugate pairs of state variable $(q^i, p_i)$ (i.e. a point in the cotangent bundle $T^*\mathcal{Q}$) and the number of possible initial conditions for each pair is given by $dq^i \wedge dp_i$ (i.e. the canonical symplectic form $\omega$). During evolution, the distribution must be mapped point by point which leads to Hamiltonian mechanics (i.e. a symplectomorphism on $T^*\mathcal{Q}$).

In section \ref{sec:relativity} we will show that a relativistic version of Hamilton's equations arises naturally when we consider time dependent laws of evolution for distributions of classical material. This also gives the notion of deterministic and reversible evolution a characterization that is independent of any transformation that mixes time and state variables $(q^i, p_i)$.

In section \ref{sec:Lagrangian} we will introduce the assumption of kinematic equivalence, studying the motion in space is equivalent to studying the evolution of the state. Under that assumption there is a link between the state variables $(q^i, p_i)$ and the kinematic variable $(x^i, u^i)$ (i.e. position and velocity). This link will allow us to express the number of possible initial conditions for d.o.f. in terms of $dx^i$ and $du^i$ (i.e. a metric tensor $g$). Such expression allows to constrain the Hamiltonian to the one for massive particles under scalar potential forces. Such assumption also allows for a Lagrangian formulation.

In section \ref{sec:quantum_prelude} we will introduce some ideas in classical mechanics to more gently transition over quantum mechanics. We will see how classical materials (i.e. distributions over phase space) already have features that are typically associated with quantum systems and we will show how to construct a classical uncertainty principle that helps to build intuition used in the later section.

In section \ref{sec:irreducibility} we will introduce the notion of an irreducible material, one for which studying the whole distribution does not tell us anything about the motion of its infinitesimal parts, the fragment. A unit amount of such material will correspond to a quantum particle. As this means that the states of such material are invariant under permutations of the configurations for the fragments, we will show that the state space of an irreducible material is a complex vector space. As we will still be able to tell how similar or different two states are, this will gives an inner product to the state space. Such comparisons will be preserved by deterministic and reversible evolution which leads to Schroedinger equation (i.e. a unitary transformation).

In section \ref{sec:discussion} we will discuss of how this work could be extended in different directions.

@@

The work is organized into:
\begin{description}
  \item[Assumptions] these characterize the physical system we are studying and constitute the premise of our discussion. A \textbf{rationale} follows each assumption, which uses physical and sometimes philosophical arguments to motivate why (or why not) such an assumption makes sense (in a particular case).
  \item[Propositions] these characterize the mathematical properties of the physical concepts we are studying. A \textbf{justification} follows each proposition to show why such characterization follows from the physical properties of the system. While the style of a justification is formal, it is not a pure mathematical proof as it necessarily has to mix physical and mathematical arguments. Before or after each proposition there is a \textbf{discussion} that conceptually explains the motivation and the result in a less formal, more approachable way.

  Justifications and discussions are designed to be independent to allow two separate types of reading style: one focused on higher level concepts and one focused on making sure the details follow. This inevitably leads to some repetition.

@@

\section{On studying a physical system}
\label{sec:fundamental_model}

@@

\section{States and state space}
\label{sec:dynamical_systems}

@@

\section{Time dependent evolution}
\label{sec:relativity}

@@

\section{Kinematics and minimal materials}
\label{sec:Lagrangian}

@@

\section{Prelude to quantum}
\label{sec:quantum_prelude}

@@

\section{Irreducible systems}
\label{sec:irreducibility}

@@

\section{Discussion and further work}
\label{sec:discussion}

@@

We have seen that much of fundamental physics can be derived from a few simple physical assumption (and lots of hard work). In doing so we obtained a more unified picture of the different branches of fundamental physics centered around the idea of deterministic and reversible evolution of distributions of a homogeneous material. The main physical concepts can be summarized by the following points:

@@

\item A state space is not just a collection of states for the system under study but it must also capture physical properties such as physical distinguishability (represented by a topology), the ability to count states (represented by a measure, symplectic form or metric tensor) or decomposability (represented by a vector space structure).

@@

\item Classical Hamiltonian mechanics describes the evolution of a homogeneous material whose state is equivalent to the states of all its infinitesimal parts, the classical particles. The state space for classical particles is a symplectic manifold $(T^*\mathcal{Q}, \omega)$ as this allows coordinate invariant density distributions; deterministic and reversible evolution for such state space is a symplectomorphism (i.e. a canonical transformation).

\item Assuming classical particle states are fully characterized by their trajectory in space, the states also evolve according to Lagrangian mechanics and the motion is that of a massive particle under scalar/vector potential forces.

@@

\item Quantum (Hamiltonian) mechanics describes the evolution of a homogeneous material for which the state can only be reduced up to finite parts, the quantum particles, whose internal motion is undetermined. The state space for quantum particles is an inner vector product space $(\Psi, \langle \cdot , \cdot \rangle)$; deterministic and reversible evolution for such state space is a unitary transformation.
\end{itemize}

While we are sure that some mathematical details can be improved, we feel that the overall ``conceptual castle" is solid. The starting concepts are simple, the conceptual mapping of those concepts into math is often straight forward and the intuitive ideas that emerge help connecting different branches of math and physics. This is unlikely a matter of simple coincidence.

While the scope of this work is limited to re-founding and re-understanding physical theories already experimentally established, it may also has the potential to provide new insights and ideas on how to address some of the open problems in the foundations of physics.
